---
title: "Getting Started with allowayai"
author: "Ian Alloway"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with allowayai}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4
)
```

## Overview

`allowayai` is a utility toolkit for data science, sports analytics, and machine
learning workflows. It was built to solve the recurring problems that come up when
you are building predictive models for sports betting markets: How do you size
your bets correctly? How do you track model performance over time? How do you
quickly inspect what your data frames actually contain?

This vignette walks through all eight functions in the package using the bundled
`nfl_lines` dataset.

---

## Installation

```r
# Install from GitHub
devtools::install_github("ianalloway/allowayai")
```

```{r}
library(allowayai)
```

---

## The `nfl_lines` Dataset

The package ships with a sample dataset of 100 simulated NFL game betting lines.
It is an **external dataset** bundled with the package and loaded lazily.

```{r}
data(nfl_lines)
head(nfl_lines)
```

The five columns are:

- `game_id`: unique game identifier
- `team`: the team being bet on
- `model_prob`: the model's predicted win probability
- `closing_odds`: the bookmaker's closing decimal odds
- `outcome`: actual result (1 = win, 0 = loss)

---

## 1. Implied Probability Conversion

Before applying your model, you need to know what the market thinks. The
`implied_prob()` function converts decimal odds to implied probabilities.

```{r}
# Single conversion
implied_prob(2.0)

# Remove the bookmaker's vig from a two-outcome market
implied_prob(odds = 1.91, remove_vig = TRUE, odds_b = 1.91)
```

You can apply it across the whole dataset:

```{r}
nfl_lines$implied <- implied_prob(nfl_lines$closing_odds)
head(nfl_lines[, c("model_prob", "closing_odds", "implied")])
```

---

## 2. Edge Summary (S3 Class)

The `edge_summary()` function creates an S3 object of class `edge_summary` that
captures the distribution of edges (model probability minus implied probability)
across all games.

```{r}
es <- edge_summary(
  model_probs = nfl_lines$model_prob,
  market_odds = nfl_lines$closing_odds,
  outcomes    = nfl_lines$outcome,
  label       = "NFL Sample Model"
)
print(es)
```

The `summary()` method provides a statistical breakdown of the edge distribution:

```{r}
summary(es)
```

---

## 3. Plot the Edge Distribution

The `plot_edge()` function creates a `ggplot2` histogram of the edge distribution.
Positive edges (blue) indicate games where the model sees value; negative edges
(red) indicate games where the market is more confident than the model.

```{r}
plot_edge(es)
```

---

## 4. Kelly Criterion Bet Sizing (S3 Class)

Once you know you have an edge, `calc_kelly()` tells you how much to bet.
It returns an S3 object of class `kelly_bet`.

```{r}
# 60% win probability at +100 odds (2.0 decimal), Half-Kelly
bet <- calc_kelly(prob = 0.60, odds = 2.0, fraction = 0.5)
print(bet)
```

The `print.kelly_bet()` S3 method formats the output to show your edge, the full
Kelly percentage, and your recommended fractional bet size.

---

## 5. Sports Model Tracker (S4 Class)

The `SportsModel` S4 class provides strict, validated storage for backtest
results. The `make_sports_model()` constructor enforces that accuracy is between
0 and 1 and that the game count is positive.

```{r}
model <- make_sports_model(
  accuracy     = 0.683,
  roi          = 0.124,
  games_tested = 6000
)
model
```

Attempting to create an invalid object raises an error immediately:

```{r, error = TRUE}
# This will fail — accuracy cannot exceed 1
make_sports_model(accuracy = 1.5, roi = 0.10, games_tested = 100)
```

---

## 6. Combined Model Summary

The `model_summary()` function bridges the S3 and S4 systems. It accepts a
`SportsModel` S4 object and an optional `edge_summary` S3 object and prints a
formatted report.

```{r}
model_summary(model, edge = es)
```

---

## 7. Data Frame Type Inspection

The `summarize_types()` utility returns a clean data frame showing the class,
base type, and S4 status of every column. It is useful when working with messy
data pulled from external APIs.

```{r}
summarize_types(nfl_lines)
```

---

## Summary of Classes and Methods

| Object | Class System | Constructor | Methods |
|---|---|---|---|
| `kelly_bet` | S3 | `calc_kelly()` | `print.kelly_bet()` |
| `edge_summary` | S3 | `edge_summary()` | `print.edge_summary()`, `summary.edge_summary()` |
| `SportsModel` | S4 | `make_sports_model()` | `show()` |

---

## License

`allowayai` is released under the **MIT License**. See `LICENSE` for details.

---

*Ian Alloway — [ianalloway.xyz](https://ianalloway.xyz) | [AllowayAI on Substack](https://allowayai.substack.com)*
