---
title: "Getting Started with allowayai"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with allowayai}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(allowayai)
```

## Introduction

Welcome to `allowayai`, a utility toolkit for data science, sports analytics, and R object exploration. This package was built for the final project and encapsulates several workflows I frequently write about on my Substack.

## Fractional Kelly Criterion

When building sports betting models, having an edge isn't enough. You need to size your bets correctly. The `calc_kelly()` function helps you do this using the Kelly Criterion, with built-in fractional sizing to reduce variance.

```{r}
# 60% model probability, +100 odds (2.0 decimal), Half-Kelly
bet <- calc_kelly(prob = 0.60, odds = 2.0, fraction = 0.5)
print(bet)
```

## S4 Classes for Model Tracking

R's S3 system is great, but sometimes you need the strictness of S4. We include an S4 class `SportsModel` to track backtest performance safely.

```{r}
model_perf <- make_sports_model(accuracy = 0.683, roi = 0.124, games_tested = 6000)
model_perf
```

## Data Type Summaries

Ever wonder what's really going on under the hood of your data frames? `summarize_types()` breaks it down.

```{r}
summarize_types(head(mtcars))
```
